### 2023-05-06 submission/manuscript.tex
Reorganize figures

McMurdie and Holmes analyzed the effect of rarefying and other
approaches on clustering accuracy using what they called ``Simulation
A'' in their Figure 2A and elsewhere in their paper. In Simulation A,
they investigated the ability to correctly assign simulated microbiome
samples to one of two clusters representing two simulated treatment
groups. Thankfully, the R code for Simulation A was provided by the
authors in the
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
R-flavored markdown file that was published as Protocol S1 in the
original paper. I will outline the simulation strategy below and will
reference line numbers from their R-flavored markdown file with ``L'' as
a prefix.

@@

After generating the OTU counts for the 160 simulated communities, the
authors applied several normalization methods, distance calculations,
and clustering algorithms to the data. To normalize the OTU data the
original analysis either applied no normalization procedure, calculated
OTU relative abundances, subsampled the data, performed variance
stabilization using the DESeq R package, or performed Upper-Quartile
log-fold change normalization using the edgeR R package. In subsampling
the data, the authors either included all of the samples or removed
samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40
percentile across all 80 samples. Subsampled data were then used as
input to calculate distances between samples using Bray-Curtis,
Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as
implemented in the Phyloseq R package, Poisson distance as implemented
in the PoiClaClu R package, and top-mean squared difference as
implemented in the edgeR R package. Un-normalized data were used to
calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac
distances. Relative abundance data were used to calculate Bray-Curtis,
Unweigthed UniFrac, and Weighted UniFrac distances. DESeq variance
stabilization normalized data were used to calculate Bray-Curtis,
Euclidean, and Weighted Unifrac distances. The Upper-Quartile log-fold
change normalized data were only used to calculate top-mean squared
difference distances. The resulting distance matrices were used to
cluster the 80 samples into one of two clusters using partitioning
around the medoid (PAM), K-means clustering, and hierarchical
clustering. Although data for all three methods were presented in the
supplementary Protocol S1, only the PAM data are presented in the main
manuscript. The accuracy of the clustering assignments was quantified as
the fraction of the 80 samples that were assigned to the correct
cluster. Because some of the subsampling conditions removed samples
those were counted as mis-clustered samples yielding minimum accuracies
below 50\%.

\hypertarget{critique-of-the-original-simulation-design}{%
\subsubsection{Critique of the original simulation
design}\label{critique-of-the-original-simulation-design}}

@@

These points will serve as an outline for the Results section. After
replicating the original simulations, these points will be evaluated to
reassess whether subsampling or rarefaction are ``inadmissible''.

@@

Before assessing the impact of the points I critiqued above, I attempted
to replicate the results shown in Figures 4 and 5 of the original paper
using the authors' original code. I created a Conda environment that
used the R version and package versions that were as close as possible
to those used in the original paper. Because of the slight differences
in packages, it was necessary to apply several patches to the original
R-flavored markdown file to render document. I was able to generate a
figure similar to that presented as Figures 4 and 5 of the original
paper. My results are shown in Figures S2
(\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_3.pdf}) and
S3 (\textbf{norarefy-source/simulation-cluster-accuracy/Figure\_4.pdf})
of this paper, respectively. The differences in results are likely due
to differences in software versions and operating systems. It is also
worth noting that the published versions of the two figures differ from
those included in Protocol S1 within the rendered html file
(\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
and that the figure numbers are one higher in the paper than those
generated by the R-flavored markdown file (i.e., Protocol S1 has Figures
3 and 4 corresponding to the published Figures 4 and 5). Regardless of
the differences, my results were qualitatively similar to the originals.

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reports the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in the original Figures 4 and 5 (e.g.~the
Bray-Curtis distances calculated on the DESeqVS normalized data). A
better approach would have been to use 100 replicates as this would
reduce the dependency of the results on the random number generator's
seed. By increasing the number of replicates it was also possible to
compare the probability of falsely and correctly clustering samples from
the same and different treatment groups together (see points 9-11,
below). Because the accuracies were not symmetric around the mean
accuracy values the median and 95\% confidence intervals or
intraquartile range should have been reported. To test the effect of
increasing the number of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment I used above. This was
necessary since the number of simulated conditions increased 20-fold
with the additional replicates. Such intense data processing was not
practical within a single R-flavored markdown document. Again, the
observed results were qualitiatively similar to those generated using
the single R-flavored markdown file (Figures
\textbf{pam\_subsample15\_fig\_4.pdf} and
\textbf{pam\_subsample\_fig\_5.pdf}). The increased number of
replications resulted in smoother lines and allowed me to present
empirical 95\% confidence intervals. For all analyses in the remainder
of this paper, I used 100 randomized replicates per condition.

@@

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. For more
modern datasets, it would be reasonable to expect a median number of
sequences larger than 10,000 (see Table 1 of {[}Singleton Paper
XXXXXXXX{]}). Therefore, I included an additional median depth of
sampling value of 50,000 sequences with the original four median
sequencing sampling depths (i.e., 1,000, 2,000, 5,000, 10,000).
Additional sequencing coverage would be expected to result in more
robust distance values since there would be more information represented
in the data. Indeed, the added sampling depth showed higher accuracy
values at lower effect sizes for the combinations of normalization
methods and distance calculations (Figure S4
(\textbf{pam\_subsample15\_fig\_4.pdf})). Increased sequencing coverage
also resulted in improved clustering accuracy for lower effect sizes
when the library size minimum quantile was decreased (Figure S5
(\textbf{pam\_subsample\_fig\_5.pdf})). I will revisit the choice of the
library size minimum quantile below.

@@

Close comparison of the original Figure 4 and my version
(\textbf{pam\_subsample15\_fig\_4.pdf}) revealed one important
difference between the two plots. In the original analysis, the
accuracies for the Weighted UniFrac distances at the largest effect size
(i.e., 3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and
10,000. In my version of the analysis, the values for the same
sequencing depths were 0.88, 0.89, and 1.00, respectively. The 95\%
confidence interval for these accuracies spaned between 0.51 and 1.00.
The Bray-Curtis distances were also different by both methods at smaller
effect sizes and had wide confidence intervals. Inspection of the DESeq
normalized OTU counts revealed that the method resulted in negative
values. In fact, rendering the R-flavored markdown files in WNWN's
Protocol S1 generated warning messages when passing the DESeq normalized
counts to the Bray-Curtis calculator, which said, ``results may be
meaningless because data have negative entries in method `bray'\,''.
Although the Weighted UniFrac calculator function did not generate a
similar warning message, negative values would also result in similarly
meaningless distances. Both are due to the fact that the calculators sum
the counts of each OTU in both samples being compared. In contrast, a
Euclidean distance does not use a similar sum, but sums the square of
the difference between the OTU abundance in each sample. To assess the
prevalence of negative counts in the simulated data, I quantified the
fraction of negative values in the OTU matrix from each simulation and
counted the number of simulations where the normalized OTU table had at
least one negative value (Figure \textbf{deseq\_negative\_value.pdf}).
In general the fraction of negative OTU counts increased with effect
size, but decreased with sequencing effort. The fraction of simulations
with at least onee negative value increased with effect size and
sequencing effort. The high frequency of negative OTU counts resulted in
highly variable Bray-Curtis and Weighted UniFrac values. It is likely
that because the WNWN analysis only used 5 replicates that the large
variation in accuracies at high effect sizes was missed initially. For
the rest of this reanalysis study, I will only report results using the
DESeq-based variance stabilization normalization with the Euclidean
distance.

@@

As noted above, the original jargon that was used in WNWN was confusing
to many who conflated rarefying/subsampling with rarefaction. A more
robust analysis would have used rarefaction since it would have averaged
across random subsamplings, which individually would be unlikely to
represent the overall composition of the communities. Rather than being
guilty of ``omission of available valid data'' as claimed in WNWN,
rarefaction uses all of the available data. To compare subsampling and
rarefaction, I removed the 15\% of samples with the lowest number of
sequences for each of the 100 simulated datasets and compared the
clustering accuracies from a single subsampling to rarefaction with 100
randomizations. This analysis revealed two benefits of rarefaction.
First, the median distances generated by rarefaction was always at least
as large as those from a single subsample (Figure
\textbf{subsample\_rarefaction\_compare.pdf}). The difference was most
pronounced for smaller average library sizes and at smaller effect
sizes. The Unweighted UniFrac distances were most impacted by the use of
rarefaction over subsampling. Second, the intraquartile ranges for the
distances generated by rarefaction were generally smaller than those by
subsampling and showed similar trends to the difference in the median
distances (Figure \textbf{subsample\_rarefaction\_compare.pdf}). The
intraquartile ranges for Bray-Curtis, Euclidean, and Unweighted UniFrac
distances were actually larger by rarefaction than by subsampling at
small effect sizes and average library sizes; however at larger values
the subsampling intraquartile range was larger than by rarefaction for
these distance calculations. Because rarefaction incorporates more of
the data and generally performed better than subsampling, the remainder
of this analysis will report results using rarefaction rather than by
subsampling, except when noted.

@@

The clustering accuracy measurements in the body of the manuscript were
determined using PAM-based clusters while Protocol S1 also includes
K-means and hierarchichal clustering. Although the data were not
displayed in a manner that lent itself to direct comparison in Protocol
S1, close inspection of the rendered figures suggested that PAM may not
have been the optimal choice in all situations. Rather, K-means
clustering appeared to perform better in many simulations. Because the
accuracies were the smallest at lower effect sizes, I focused my
comparison at the effect size of 1.15. For each set of 100 replicated
simulated datasets, I compared the clustering accuracy across clustering
methods to see how often each clustering method resulted in the highest
accuracy (Figure \textbf{compare\_cluster\_methods.pdf}). Indeed,
K-means clustering performed better than the other methods. Among all
combinations of normalization methods, distance calculations, and read
depths, PAM clustering resulted in clustering accuracies as good or
better than the other methods in 49.92\% of the randomizations (Figure
\textbf{compare\_cluster\_methods.pdf}). K-means clustering was at least
as good as the other methods in 74.39\% of the randomizations. HClust
was at least as good as the other methods for 44.32\% of the
randomizations. I specifically compared the clustering accuracies using
rarefaction for each of the distance calculations methods using PAM and
K-means clustering. Among the 30 combinations of distance calculations
and read depths, K-means performed better than PAM in 29 cases with PAM
doing better in the 1 other case (i.e., calculating distances with
Euclidean using 10,000 sequences). When using subsampled data, K-means
clustering performed better than PAM in each case. Because K-means
clustering did so much better than PAM clustering in the simulated
conditions, I will use K-means clustering for the remainder of this
study.

@@

In WNWN, the authors quantified the tradeoff between median sequencing
depth, the number of samples removed below the threshold, and clustering
accuracy (original Figure 5, my Figure S5). Although the optimal
threshold varied by distance metric, normalization method, and
sequencing depth, they removed samples whose number of sequences was
less than the 15th percentile (L404-419). They acknowledged that this
screening step, which was only used with subsampling, would decrease
clustering accuracy putting it at a relative disadvantage to the other
methods (page 5, column 1, last paragraph). Therefore, it was not
surprising that the peak clustering accuracy for their subsampled data
was at 85\%. Because the true best result would not known \emph{a
priori} in an actual microbiome study, it would be impossible for
researchers to conduct a sensitivity analysis comparing the tradeoffs
between sequencing depth, sample number, and clustering accuracy to
select a sampling depth for their analysis. The differences in
clustering accuracy between subsampling and rarefaction and using PAM
and K-means clustering indicated that it was necessary to reassess the
tradeoff between the library size minimum quantile and clustering
accuracy. When using rarefaction, K-means clustering, and only
considering conditions with 2,000 or more sequences, there was not a
condition where setting a higher threshold resulted in a better accuracy
than using all of the samples
(\textbf{kmeans\_rarefaction\_fig\_5.pdf}). These results showed that
for modern sequencing depths, using the full datasets with rarefaction
and K-means clustering resulted in accuracies that were typically better
than those observed when removing the smallest 15\% of the samples from
each simulated dataset. When the original Figure 4 was recast with these
approaches, rarefaction performed at least as well as any of the other
tranformations with each distance calculation, except when used with the
Poisson distance (Figure \textbf{kmeans\_rarefaction15\_fig\_4.pdf}). It
is worth noting that at the largest effect sizes, K-means clustering did
not perform as well as PAM for some combinations of normalization method
and distance calculation (compare
\textbf{kmeans\_rarefaction15\_fig\_4.pdf} and
\textbf{pam\_subsample00\_fig\_4.pdf}); however, those combinations that
performed worse by K-means were not as good as rarefaction or
subsampling by either clustering method.

@@

As described above, the sequencing depths used in the 26 GlobalPatterns
datasets were used as the distribution to create sequencing depths for
the 80 samples that were generated in each simulation. The
GlobalPatterns datasets had a mean of 1085256.8 sequences and a median
of 1,106,849 sequences per dataset. The datasets ranged in sequencing
depth between 58,688 and 2,357,181 sequences for a 40.16-fold
difference. Rather than representing a typically observed distribution
of sequencing depths that would be skewed right, the sampling
distribution was normally distributed (Shapiro-Wilk test of normality,
P=0.57) (Figure \textbf{\emph{distribution\_shape.pdf}}). From these
simulations it is unclear how sensitive the various normalizations and
distance calculations were to a skewed distribution. A second limitation
of this sampling distribution is that it only contained 26 unique
sampling depths such that each sampling depth would have been re-used an
average of 3.08 times in each simulation. Yet, it is unlikely for a real
sequence collection to have duplicate sequencing depths. To reassess the
WNWN results in the context of a more typical distribution of sample
sizes, I created a new set of simulations to test the effect of the
shape of the distribution on the results. I created a simple sequencing
depth distribution where there were 80 depths logarithmically
distributed between the minimum and maximum sequencing depths of the
GlobalPatterns dataset (Figure \textbf{\emph{distribution\_shape.pdf}}).
The median of this distribution was 372,040 and the mean was 629824.8.
When I regenerated the original Figures 4 and 5 using the
log-distributed sequencing effort distribution, the differences in
normalization methods were more apparent (Figure
\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_log.pdf}}). For each of the
distance calculators, rarefaction to the size of the smallest dataset
yielded accuracies that were at least as good as the other methods
across effect sizes and median sequencing depths. The difference was
most pronounced at smaller effect sizes and sequencing depths. When
comparing the performance of rarefaction across distance calculators for
different effect sizes, sequencing depths and size of smallest sample
(\textbf{\emph{fig\_5\_kmeans\_rarefaction\_log.pdf}}), the accuracies I
observed using the log-distributed sample sizes was at least as good as
those obtained using the GlobalPatterns-based distribution
\textbf{\emph{fig\_4\_kmeans\_rarefaction00\_a.pdf}}. The issue of the
number of samples in a study and the distribution of their sequencing
depths in the context of controlling for uneven sampling effort is
explored in far greater detail in an another analysis using sequencing
depths observed in actual biological samples {[}XXXXXX{]}.

@@

McMurdie and Holmes were emphatic that ``\textbf{rarefying biological
count data is statistically inadmissible} because it requires the
omission of available valid data'' (emphasis in original). Thus it is
strange that they argue against removing data when
rarefying/subsampling, but accept removing rare and low-prevalence OTUs
prior to normalizing their counts. This practice has become common in
microbiome studies and is the standard approach in tools such as dada2,
unoise, and deblur {[}XXXXXXXX{]}. However, my previous work has shown
that rare sequences from a poorly sequenced sample often appear in more
deeply sequened samples suggesting that they are not necessarily
artifacts. Furthermore, removing rare sequences alters the structure of
communities and has undesirable effects on downstream analyses
{[}XXXXXXXX{]}. Although my previous work does an extensive analysis of
the effects of removing rare sequences, I wanted to explore the effect
of filtering in the context of the WNWN simulation framework. For each
of the filtered and non-filtered OTU tables I calcualted the absolute
value of the difference in accuracy between each distance calculation
following the normalization procedure (Figure
\textbf{compare\_filter\_accuracy.pdf}). With the exception of the
Weighted UniFrac distances, each of the distance calculations and
normalization procedures were sensitive to the filtering. The
rarefaction data tended to be sensitive to filtering at small effect
sizes. Distances generated using raw counts, DESeq variance
stabilization, and Upper Quartile Log Fold Change tended to be more
sensitive to filtering at larger effect sizes. When using relative
abundance data, Bray-Curtis distances were sensitive to filtering at
small effect sizes and Unweighted UniFrac distances were sensitive at
large effect sizes. These trends appeared to be driven by the dependence
of the distance calculation on low abundance taxa. More surprising than
the effect of filtering on the mean absolute difference in clustering
accuracy was the wide variation in accuracies at each effect size. Among
the different normalization methods, the accuracies calculated using
rarefaction had the narrowest 95\% confidence interval for all distance
calculations except for calculating Unweighted UniFrac distances. For
these distances, using relative abundances had the narrowest range at
small effect sizes; rarefaction had the narrowest range at larger effect
sizes. Again, these trends appeared to be driven by the dependence on
low frequency taxa in Unweighted UniFrac, which is dependent on the
presence or absence of taxa rather than their abundance. Given my
previous work and the large variation caused by removing rare taxa, OTU
filtering should not be performed in microbiome analyses.

>>>>

McMurdie and Holmes analyzed the effect of rarefying and other
approaches on clustering accuracy using what they called ``Simulation
A'' in their Figure 2A and elsewhere in their paper. In Simulation A,
they investigated the ability to correctly assign simulated microbiome
samples to one of two clusters representing two simulated treatment
groups. Thankfully, the R code for Simulation A was provided by the
authors in the
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
R-flavored markdown file that was published as Protocol S1 in WNWN. I
will outline the simulation strategy below and will reference line
numbers from their R-flavored markdown file with ``L'' as a prefix.

@@

After generating the OTU counts for the 160 simulated communities, the
authors applied several normalization methods, distance calculations,
and clustering algorithms to the data. To normalize the OTU data the
WNWN analysis either applied no normalization procedure, calculated OTU
relative abundances, subsampled the data, performed variance
stabilization using the DESeq R package, or performed Upper-Quartile
log-fold change normalization using the edgeR R package. In subsampling
the data, the authors either included all of the samples or removed
samples whose sequencing depth fell below the 5, 10, 15, 20, 25, and 40
percentile across all 80 samples. Subsampled data were then used as
input to calculate distances between samples using Bray-Curtis,
Euclidean, Unweighted UniFrac, and Weighted Unifrac distances as
implemented in the Phyloseq R package, Poisson distance as implemented
in the PoiClaClu R package, and top-mean squared difference as
implemented in the edgeR R package. Un-normalized data were used to
calculate Bray-Curtis, Euclidean, Poisson, and Weighted UniFrac
distances. Relative abundance data were used to calculate Bray-Curtis,
Unweigthed UniFrac, and Weighted UniFrac distances. DESeq variance
stabilization normalized data were used to calculate Bray-Curtis,
Euclidean, and Weighted Unifrac distances. The Upper-Quartile log-fold
change normalized data were only used to calculate top-mean squared
difference distances. The resulting distance matrices were used to
cluster the 80 samples into one of two clusters using partitioning
around the medoid (PAM), K-means clustering, and hierarchical
clustering. Although data for all three methods were presented in the
supplementary Protocol S1, only the PAM data are presented in the main
manuscript. The accuracy of the clustering assignments was quantified as
the fraction of the 80 samples that were assigned to the correct
cluster. Because some of the subsampling conditions removed samples
those were counted as mis-clustered samples yielding minimum accuracies
below 50\%.

\hypertarget{critique-of-the-wnwn-simulation-design}{%
\subsubsection{Critique of the WNWN simulation
design}\label{critique-of-the-wnwn-simulation-design}}

@@

These points will serve as an outline for the Results section. After
replicating the WNWN simulations, these points will be evaluated to
reassess whether subsampling or rarefaction are ``inadmissible''.

@@

Before assessing the impact of the points I critiqued above, I attempted
to replicate the results shown in Figures 4 and 5 of WNWN using the
authors' code. I created a Conda environment that used the R version and
package versions that were as close as possible to those used in WNWN.
Because of the slight differences in packages, it was necessary to apply
several patches to WNWN's R-flavored markdown file to render document. I
was able to generate figures similar to those presented as WNWN's
Figures 4 and 5. My results are shown in Figures S1 and S2
(\textbf{Figure s1\_figure\_3.pdf}/\textbf{Figure s1\_figure\_4.pdf}),
respectively. The differences in results are likely due to differences
in software versions and operating systems. It is also worth noting that
the published versions of the two figures differ from those included in
Protocol S1 within the rendered html file
(\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.html})
and that the figure numbers are one higher in the paper than those
generated by the R-flavored markdown file (i.e., Protocol S1 labels the
figures as Figures 3 and 4 corresponding to the published Figures 4 and
5). Regardless of the differences, my results were qualitatively similar
to that provided in WNWN's Protocol S1.

@@

Each simulated condition was replicated 5 times in WNWN and the paper
reports the mean and standard deviation of the replicate clustering
accuracies. The relatively small number of replicates accounts for the
jerkiness of the lines in WNWN's Figures 4 and 5 (e.g.~the Bray-Curtis
distances calculated on the DESeqVS normalized data). A better approach
would have been to use 100 replicates as this would reduce the
dependency of the results on the random number generator's seed. By
increasing the number of replicates it was also possible to compare the
probability of falsely and correctly clustering samples from the same
and different treatment groups together (see points 9-11, below).
Because the accuracies were not symmetric around the mean accuracy
values the median and 95\% confidence intervals or intraquartile range
should have been reported. To test the effect of increasing the number
of replicates, I pulled apart the code in
\texttt{simulation-cluster-accuracy/simulation-cluster-accuracy-server.Rmd}
into individual R and bash scripts that were executed using a Snakemake
workflow with the same Conda environment I used above. This was
necessary since the number of simulated conditions increased 20-fold
with the additional replicates. Such intense data processing was not
practical within a single R-flavored markdown document. Again, the
observed results were qualitiatively similar to those generated using
the single R-flavored markdown file (Figures S3 and S4
\textbf{pam\_subsample15\_fig\_4.pdf}/\textbf{pam\_subsample\_fig\_5.pdf}).
The increased number of replications resulted in smoother lines and
allowed me to present empirical 95\% confidence intervals. For all
analyses in the remainder of this paper, I used 100 randomized
replicates per condition.

@@

In the 10 years since WNWN was published, sequencing technology has
advanced and sequence collections have grown considerably. For more
modern datasets, it would be reasonable to expect a median number of
sequences larger than 10,000 (see Table 1 of {[}Singleton Paper
XXXXXXXX{]}). Therefore, I included an additional median depth of
sampling value of 50,000 sequences with WNWN's four median sequencing
sampling depths (i.e., 1,000, 2,000, 5,000, 10,000). Additional
sequencing coverage would be expected to result in more robust distance
values since there would be more information represented in the data.
Indeed, the added sampling depth showed higher accuracy values at lower
effect sizes for the combinations of normalization methods and distance
calculations (Figure S3 \textbf{fig4\_pam\_subsample15\_gp\_ds.pdf}).
Increased sequencing coverage also resulted in improved clustering
accuracy for lower effect sizes when the library size minimum quantile
was decreased (Figure S4 \textbf{fig5\_pam\_subsample\_gp.pdf}). I will
revisit the choice of the library size minimum quantile below.

@@

Close comparison of WNWN's Figure 4 and my version (Figure S3
\textbf{fig4\_pam\_subsample15\_gp\_ds.pdf}) revealed one distinct
difference between the two plots. In the WNWN analysis, the accuracies
for the Weighted UniFrac distances at the largest effect size (i.e.,
3.5) were 1.0 for median sequencing depths of 1,000, 2,000, and 10,000.
In my version of the analysis, the values for the same sequencing depths
were 0.88, 0.89, and 1.00, respectively. The 95\% confidence interval
for these accuracies spaned between 0.51 and 1.00. The Bray-Curtis
distances were also different by both methods at smaller effect sizes
and had wide confidence intervals. Inspection of the DESeq normalized
OTU counts revealed that the method resulted in negative values. In
fact, rendering the R-flavored markdown files in WNWN's Protocol S1
generated warning messages when passing the DESeq normalized counts to
the Bray-Curtis calculator, which said, ``results may be meaningless
because data have negative entries in method `bray'\,''. Although the
Weighted UniFrac calculator function did not generate a similar warning
message, negative values would also result in similarly meaningless
distances. Both are due to the fact that the calculators sum the counts
of each OTU in both samples being compared. In contrast, a Euclidean
distance does not use a similar sum, but sums the square of the
difference between the OTU abundance in each sample. To assess the
prevalence of negative counts in the simulated data, I quantified the
fraction of negative values in the OTU matrix from each simulation and
counted the number of simulations where the normalized OTU table had at
least one negative value (Figure S5
\textbf{deseq\_negative\_value.pdf}). In general the fraction of
negative OTU counts increased with effect size, but decreased with
sequencing effort. The fraction of simulations with at least onee
negative value increased with effect size and sequencing effort. The
high frequency of negative OTU counts resulted in highly variable
Bray-Curtis and Weighted UniFrac values. It is likely that because the
WNWN analysis only used 5 replicates that the large variation in
accuracies at high effect sizes was missed initially. For the rest of
this reanalysis study, I will only report results using the DESeq-based
variance stabilization normalization with the Euclidean distance.

@@

As noted above, the jargon used in WNWN was confusing to many who
conflated rarefying/subsampling with rarefaction. A more robust analysis
would have used rarefaction since it would have averaged across random
subsamplings, which individually would be unlikely to represent the
overall composition of the communities. Rather than being guilty of
``omission of available valid data'' as claimed in WNWN, rarefaction
uses all of the available data. To compare subsampling and rarefaction,
I removed the 15\% of samples with the lowest number of sequences for
each of the 100 simulated datasets and compared the clustering
accuracies from a single subsampling to rarefaction with 100
randomizations. This analysis revealed two benefits of rarefaction.
First, the median distances generated by rarefaction was always at least
as large as those from a single subsample (Figure 1
\textbf{subsample\_rarefaction\_compare.pdf}). The difference was most
pronounced for smaller average library sizes and at smaller effect
sizes. The Unweighted UniFrac distances were most impacted by the use of
rarefaction over subsampling. Second, the intraquartile ranges for the
distances generated by rarefaction were generally smaller than those by
subsampling and showed similar trends to the difference in the median
distances (Figure 1 \textbf{subsample\_rarefaction\_compare.pdf}). The
intraquartile ranges for Bray-Curtis, Euclidean, and Unweighted UniFrac
distances were actually larger by rarefaction than by subsampling at
small effect sizes and average library sizes; however at larger values
the subsampling intraquartile range was larger than by rarefaction for
these distance calculations. Because rarefaction incorporates more of
the data and generally performed better than subsampling, the remainder
of this analysis will report results using rarefaction rather than by
subsampling, except when noted.

@@

The clustering accuracy measurements in the body of the manuscript were
determined using PAM-based clusters while Protocol S1 also includes
K-means and hierarchichal clustering. Although the data were not
displayed in a manner that lent itself to direct comparison in Protocol
S1, close inspection of the rendered figures suggested that PAM may not
have been the optimal choice in all situations. Rather, K-means
clustering appeared to perform better in many simulations. Because the
accuracies were the smallest at lower effect sizes, I focused my
comparison at the effect size of 1.15. For each set of 100 replicated
simulated datasets, I compared the clustering accuracy across clustering
methods to see how often each clustering method resulted in the highest
accuracy (Figure 2 \textbf{compare\_cluster\_methods.pdf}). Indeed,
K-means clustering performed better than the other methods. Among all
combinations of normalization methods, distance calculations, and read
depths, PAM clustering resulted in clustering accuracies as good or
better than the other methods in 49.92\% of the randomizations (Figure 2
\textbf{compare\_cluster\_methods.pdf}). K-means clustering was at least
as good as the other methods in 74.39\% of the randomizations. HClust
was at least as good as the other methods for 44.32\% of the
randomizations. I specifically compared the clustering accuracies using
rarefaction for each of the distance calculations methods using PAM and
K-means clustering. Among the 30 combinations of distance calculations
and read depths, K-means performed better than PAM in 29 cases with PAM
doing better in the 1 other case (i.e., calculating distances with
Euclidean using 10,000 sequences). When using subsampled data, K-means
clustering performed better than PAM in each case. Because K-means
clustering did so much better than PAM clustering in the simulated
conditions, I will use K-means clustering for the remainder of this
study.

@@

In WNWN, the authors quantified the tradeoff between median sequencing
depth, the number of samples removed below the threshold, and clustering
accuracy (WNWN's Figure 5, my Figure S4
\textbf{fig5\_pam\_subsample\_gp.pdf}). Although the optimal threshold
varied by distance metric, normalization method, and sequencing depth,
they removed samples whose number of sequences was less than the 15th
percentile (L404-419). They acknowledged that this screening step, which
was only used with subsampling, would decrease clustering accuracy
putting it at a relative disadvantage to the other methods (page 5,
column 1, last paragraph). Therefore, it was not surprising that the
peak clustering accuracy for their subsampled data was at 85\%. Because
the true best result would not known \emph{a priori} in an actual
microbiome study, it would be impossible for researchers to conduct a
sensitivity analysis comparing the tradeoffs between sequencing depth,
sample number, and clustering accuracy to select a sampling depth for
their analysis. The differences in clustering accuracy between
subsampling and rarefaction and using PAM and K-means clustering
indicated that it was necessary to reassess the tradeoff between the
library size minimum quantile and clustering accuracy. When using
rarefaction, K-means clustering, and only considering conditions with
2,000 or more sequences, there was not a condition where setting a
higher threshold resulted in a better accuracy than using all of the
samples (Figure 3 \textbf{kmeans\_rarefaction\_fig\_5.pdf}). These
results showed that for modern sequencing depths, using the full
datasets with rarefaction and K-means clustering resulted in accuracies
that were typically better than those observed when removing the
smallest 15\% of the samples from each simulated dataset. When the WNWN
Figure 4 was recast with these approaches, rarefaction performed at
least as well as any of the other tranformations with each distance
calculation, except when used with the Poisson distance (Figure 4
\textbf{fig4\_kmeans\_rarefaction00\_gp\_nds.pdf}). \textbf{It is worth
noting that at the largest effect sizes, K-means clustering did not
perform as well as PAM for some combinations of normalization method and
distance calculation; however, those combinations that performed worse
by K-means were not as good as rarefaction or subsampling by either
clustering method.}

@@

As described above, the sequencing depths used in the 26 GlobalPatterns
datasets were used as the distribution to create sequencing depths for
the 80 samples that were generated in each simulation. The
GlobalPatterns datasets had a mean of 1085256.8 sequences and a median
of 1,106,849 sequences per dataset (Figure 5
\textbf{distribution\_shape.pdf}). The datasets ranged in sequencing
depth between 58,688 and 2,357,181 sequences for a 40.16-fold
difference. Rather than representing a typically observed distribution
of sequencing depths that would be skewed right, the sampling
distribution was normally distributed (Shapiro-Wilk test of normality,
P=0.57). From these simulations it is unclear how sensitive the various
normalizations and distance calculations were to a skewed distribution.
A second limitation of this sampling distribution is that it only
contained 26 unique sampling depths such that each sampling depth would
have been re-used an average of 3.08 times in each simulation. Yet, it
is unlikely for a real sequence collection to have duplicate sequencing
depths. To reassess the WNWN results in the context of a more typical
distribution of sample sizes, I created a new set of simulations to test
the effect of the shape of the distribution on the results. I created a
simple sequencing depth distribution where there were 80 depths
logarithmically distributed between the minimum and maximum sequencing
depths of the GlobalPatterns dataset (Figure 5
\textbf{distribution\_shape.pdf}). The median of this distribution was
372,040 and the mean was 629824.8. When I regenerated the WNWN's Figures
4 and 5 using the log-distributed sequencing effort distribution, the
differences in normalization methods were more apparent (Figure 6
\textbf{fig\_4\_kmeans\_rarefaction00\_log.pdf}). For each of the
distance calculators, rarefaction to the size of the smallest dataset
yielded accuracies that were at least as good as the other methods
across effect sizes and median sequencing depths. The difference was
most pronounced at smaller effect sizes and sequencing depths. When
comparing the performance of rarefaction across distance calculators for
different effect sizes, sequencing depths and size of smallest sample
(Figure 7 \textbf{fig\_5\_kmeans\_rarefaction\_log.pdf}), the accuracies
I observed using the log-distributed sample sizes was at least as good
as those obtained using the GlobalPatterns-based distribution (Figure 4
\textbf{fig\_4\_kmeans\_rarefaction00\_gp\_nds.pdf}). The issue of the
number of samples in a study and the distribution of their sequencing
depths in the context of controlling for uneven sampling effort is
explored in far greater detail in an another analysis using sequencing
depths observed in actual biological samples {[}XXXXXX{]}.

@@

McMurdie and Holmes were emphatic that ``\textbf{rarefying biological
count data is statistically inadmissible} because it requires the
omission of available valid data'' (emphasis in original). Thus it is
strange that they argue against removing data when
rarefying/subsampling, but accept removing rare and low-prevalence OTUs
prior to normalizing their counts. This practice has become common in
microbiome studies and is the standard approach in tools such as dada2,
unoise, and deblur {[}XXXXXXXX{]}. However, my previous work has shown
that rare sequences from a poorly sequenced sample often appear in more
deeply sequened samples suggesting that they are not necessarily
artifacts. Furthermore, removing rare sequences alters the structure of
communities and has undesirable effects on downstream analyses
{[}XXXXXXXX{]}. Although my previous work does an extensive analysis of
the effects of removing rare sequences, I wanted to explore the effect
of filtering in the context of the WNWN simulation framework. For each
of the filtered and non-filtered OTU tables I calcualted the absolute
value of the difference in accuracy between each distance calculation
following the normalization procedure (Figure S6
\textbf{compare\_filter\_accuracy.pdf}). With the exception of the
Weighted UniFrac distances, each of the distance calculations and
normalization procedures were sensitive to the filtering. The
rarefaction data tended to be sensitive to filtering at small effect
sizes. Distances generated using raw counts, DESeq variance
stabilization, and Upper Quartile Log Fold Change tended to be more
sensitive to filtering at larger effect sizes. When using relative
abundance data, Bray-Curtis distances were sensitive to filtering at
small effect sizes and Unweighted UniFrac distances were sensitive at
large effect sizes. These trends appeared to be driven by the dependence
of the distance calculation on low abundance taxa. More surprising than
the effect of filtering on the mean absolute difference in clustering
accuracy was the wide variation in accuracies at each effect size. Among
the different normalization methods, the accuracies calculated using
rarefaction had the narrowest 95\% confidence interval for all distance
calculations except for calculating Unweighted UniFrac distances. For
these distances, using relative abundances had the narrowest range at
small effect sizes; rarefaction had the narrowest range at larger effect
sizes. Again, these trends appeared to be driven by the dependence on
low frequency taxa in Unweighted UniFrac, which is dependent on the
presence or absence of taxa rather than their abundance. Given my
previous work and the large variation caused by removing rare taxa, OTU
filtering should not be performed in microbiome analyses.
